Succinct representations for abstract interpretation

### 2011-10-31 papers/PLDI2012/implicitization.tex
macros pour desactiver commentaires
notations (correction des indices)

\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
% Pour les commentaires
\usepackage[utf8]{inputenc}

@@

\JH{$99990)$, ca designe bien une borne d'intervalle ouvert ? J'ai plus
l'habitude a la notation $99990[$}
\MM{Je crois que les anglophones utilisent plutôt $99990)$, mais
  d'après Wikipedia ``Interval (mathematics)'', les deux sont acceptés
  par le standard ISO}
Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
The \lstinline|x > x_old+10| test at line~6, if taken, yields $\lstinline|x_old|
\in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain
$\lstinline|x| \in (-\infty,\allowbreak 100000)$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.

@@

For a loop header $p_i$, this technique computes a disjunctive invariant
$\bigvee_{1\leq j \leq m} X_{i,j}$. We denote by $n$ the number of
distinct paths in the loop.
To perform the analysis, one chooses an integer $\delta \in [1,m]$, and
a mapping function $\sigma: [1,m] \times [1,n] \mapsto [1,m]$.
\MM{C'est moi ou le $\tau_{i,k}$ ne correspond pas exactement à la
  notation du début ? Au début, $\tau_{i,k}$ est la transition de $i$
  à $k$, alors que là, on est dans le multigraphe donc il y a
  plusieurs arrêtes entre deux points. Si je comprends bien,
  $\tau_{i,k}$ est le $k$-ième chemin partant de $i$, c'est ça ?
  Je mettrais bien ``... and for the $k$-th path starting from $i$,
  denoted $\tau_{i,k}$, the image...'' si c'est bien ça. D'ailleurs,
  faut-il forcément parler de boucle ici ? La technique ne s'applique
  pas à des chemins quelconques ?}
  For each
abstract value of the disjunctive invariant, and for each path $\tau_{i,k}$ in the loop, the
image of $X_{i,j}$ by the path is joined with
$X_{i,\sigma(j,k)}$.
Initially, the $\delta$-th abstract value contains the initial states of
$p_i$, and all other abstract values contain~$\emptyset$.

$m$, $\delta$ and $\sigma$ can be defined heuristically.
For instance, one could define $\sigma$ so that $\sigma(j,k)$ only depends on the
last transition of the path, or else construct it dynamically during the
analysis.

@@

Our improvement consists in a modification of the SMT formula we solve in
\ref{sec:guided_multigraph}.
We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
that we can easily find in the model which abstract value of the
disjunction of the source point has
to be chosen to make the invariant of the destination grow.
The resulting formula that is given to the SMT solver is defined
by $g(p_i)$.
When the formula is satisfiable, we know that the index $j$ of the starting
disjunct that has to be chosen is the one for which the associate Boolean value
$d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
$\sigma(j,k)$, thus know the index of the disjunct to join with.

@@

The \emph{Update} function can classically assign to $X_{i,\sigma(j,k)}$ the
value $X_{i,\sigma(j,k)} \widening (X_{i,\sigma(j,k)} \sqcup
\tau_{i,k}(X_{i,j}))$, or can integrate the special treatment for self loops
proposed by \citet{Monniaux_Gonnord_SAS11}, with widening/narrowing sequence or
acceleration.

\begin{algorithm}[!h]
\caption{Disjunctive invariant computation with implicit paths}\label{gulwani2}
\label{algo:disjunctive}
\begin{algorithmic}[1] 
\WHILE {true}
		\STATE $res \gets SmtSolve\left[g(p_i)\right]$
	\IF {$res = unsat$}
		\STATE \textbf{break}
	\ENDIF
	\STATE Compute the path $\tau_{i,k}$ from $res$ 
	\STATE Take $j \in \{ l | d_l = true\}$ 
	\STATE Update($X_{i,\sigma(j,k)})$
\ENDWHILE
\end{algorithmic}
\end{algorithm}

The $\sigma$ function takes as parameters the index of the starting abstract
value, and the path we focus on. If we dynamically construct this function
during the analysis, we should store the results of the $\sigma$ function into a
compact representation, such as Algebraic Decision Diagrams.
Note that $\sigma(j,k)$ will be computed only if needed.

We experimented with a heuristic
of dynamic construction of the $\sigma$ function, adapted from
\citet{DBLP:conf/pldi/GulwaniZ10}. 
\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
et qu'elle n'était pas dans Gulwani?}
\JH{Non. Gulwani fait la même chose en un peu plus complique (voir commentaire
suivant)}
We start with one single
disjunct per control point ($m=1$) and an empty map $\sigma$.
$M$ denotes an upper bound on the number of disjuncts per control point.
$\sigma(j,k)$ is then constructed on the fly only when needed.
When $\sigma(j,k)$ needs to be computed, we first compute the image of the
abstract value $X_k$ by the path indexed by $j$, and try to find an existing
disjunct of index $k'$ so that the least upper bound of the two abstract values is exactly their union (this can be tested using SMT-solving).
If such an index exists, then we set $\sigma(j,k) = k'$.
Otherwise:
\begin{itemize}[itemsep=-1\parsep,topsep=0pt]
	\item if $m < M$, we increase $m$ by $1$ and define $\sigma(j,k) = m$
	\item if $m = M$, we define $\sigma(j,k) = M$ 
		\JH{A la place, Gulwani cherche l'invariant qui a le plus de contraintes
		d'egalite en commun pour definir son $sigma$. Nous, on le met dans M, pour simplifier}
\end{itemize}

>>>>

\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
%\newcommand{\MM}[1]{}
%\newcommand{\JH}[1]{}
%\newcommand{\DM}[1]{}

% Pour les commentaires
\usepackage[utf8]{inputenc}

@@

Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
The \lstinline|x > x_old+10| test at line~6, if taken, yields $\lstinline|x_old|
\in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain
$\lstinline|x| \in (-\infty,\allowbreak 100000)$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.

@@

For each control point $p_i$, we compute a disjunctive invariant
$\bigvee_{1\leq j \leq m_i} X_{i,j}$. We denote by $n_i$ the number of
distinct paths starting at $p_i$.
To perform the analysis, one chooses an integer $\delta_i \in [1,m_i]$, and
a mapping function $\sigma_i: [1,m_i] \times [1,n_i] \mapsto [1,m_i]$.
\MM{C'est moi ou le $\tau_{i,k}$ ne correspond pas exactement à la
  notation du début ? Au début, $\tau_{i,k}$ est la transition de $i$
  à $k$, alors que là, on est dans le multigraphe donc il y a
  plusieurs arrêtes entre deux points. Si je comprends bien,
  $\tau_{i,k}$ est le $k$-ième chemin partant de $i$, c'est ça ?
  Je mettrais bien ``... and for the $k$-th path starting from $i$,
  denoted $\tau_{i,k}$, the image...'' si c'est bien ça. D'ailleurs,
  faut-il forcément parler de boucle ici ? La technique ne s'applique
  pas à des chemins quelconques ?}
  The $k$-th path starting fom $p_i$ is denoted $\tau_{i,k}$.
  The image of the $j$-th disjunct $X_{i,j}$ by the path $\tau_{i,k}$ is joined with 
  $X_{i,\sigma(j,k)}$.
Initially, the $\delta_i$-th abstract value contains the initial states of
$p_i$, and all other abstract values contain~$\emptyset$.

For each control point $p_i \in P_R$, $m_i$, $\delta_i$ and $\sigma_i$ can be defined heuristically.
For instance, one could define $\sigma_i$ so that $\sigma_i(j,k)$ only depends on the
last transition of the path, or else construct it dynamically during the
analysis.

@@

Our improvement consists in a modification of the SMT formula we solve in
\ref{sec:guided_multigraph}.
We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
that we can easily find in the model which abstract value of the
disjunction of the source point has
to be chosen to make the invariant of the destination grow.
The resulting formula that is given to the SMT solver is defined
by $g(p_i)$.
When the formula is satisfiable, we know that the index $j$ of the starting
disjunct that has to be chosen is the one for which the associate Boolean value
$d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
$\sigma_i(j,k)$, thus know the index of the disjunct to join with.

@@

The \emph{Update} function can classically assign to $X_{i,\sigma_i(j,k)}$ the
value $X_{i,\sigma_i(j,k)} \widening (X_{i,\sigma_i(j,k)} \sqcup
\tau_{i,k}(X_{i,j}))$, or can integrate the special treatment for self loops
proposed by \citet{Monniaux_Gonnord_SAS11}, with widening/narrowing sequence or
acceleration.

\begin{algorithm}[!h]
\caption{Disjunctive invariant computation with implicit paths}\label{gulwani2}
\label{algo:disjunctive}
\begin{algorithmic}[1] 
\WHILE {true}
		\STATE $res \gets SmtSolve\left[g(p_i)\right]$
	\IF {$res = unsat$}
		\STATE \textbf{break}
	\ENDIF
	\STATE Compute the path $\tau_{i,k}$ from $res$ 
	\STATE Take $j \in \{ l | d_l = true\}$ 
	\STATE Update($X_{i,\sigma_i(j,k)})$
\ENDWHILE
\end{algorithmic}
\end{algorithm}

The $\sigma_i$ functions take as parameters the index of the starting abstract
value, and the path we focus on. If we dynamically construct these functions
during the analysis, we should store the results of the $\sigma_i$ functions into a
compact representation, such as Algebraic Decision Diagrams.
Note that $\sigma_i(j,k)$ will be computed only if needed.

We experimented with a heuristic
of dynamic construction of the $\sigma_i$ functions, adapted from
\citet{DBLP:conf/pldi/GulwaniZ10}. 
\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
et qu'elle n'était pas dans Gulwani?}
\JH{Non. Gulwani fait la même chose en un peu plus complique (voir commentaire
suivant)}
We start with one single
disjunct per control point ($m_i=1$ for each $i, p_i \in P_R$) 
and empty maps $\sigma_i$.
$M$ denotes an upper bound on the number of disjuncts per control point.
$\sigma_i(j,k)$ is then constructed on the fly only when needed.
When $\sigma_i(j,k)$ needs to be computed, we first compute the image of the
abstract value $X_{i,j}$ by the path indexed by $k$, and try to find an existing
disjunct of index $j'$ so that the least upper bound of the two abstract values is exactly their union (this can be tested using SMT-solving).
If such an index exists, then we set $\sigma_i(j,k) = j'$.
Otherwise:
\begin{itemize}[itemsep=-1\parsep,topsep=0pt]
	\item if $m_i < M$, we increase $m_i$ by $1$ and define $\sigma_i(j,k) = m_i$
	\item if $m_i = M$, we define $\sigma_i(j,k) = M$ 
		\JH{A la place, Gulwani cherche l'invariant qui a le plus de contraintes
		d'egalite en commun pour definir son $sigma$. Nous, on le met dans M, pour simplifier}
\end{itemize}
